---
title: "Loan Analysis"
author: "Collins Oduor"
date: "2026-03-31"
output: pdf_document
---

# Configuration

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE) # Ensure the code blocks are shown when the markdown is knitted to the target format
```

# Imports

```{r}
required_libraries <- c("dplyr", "tidyverse", "corrr", "corrplot", "MASS")

for (required_library in required_libraries) {
  if(!(required_library %in% installed.packages())) {
    install.packages(required_library)
  }
  
  library(required_library, character.only = T)
}
```

# Loading data

```{r}
df <- read_csv(file="loan_data.csv")
```

# Exploratory Analysis
## Summary of data

```{r}
dim(df)
```

```{r}
head(df)
```

```{r}
names(df)
```

```{r}
sapply(df, anyNA)
```
## Univariate Analysis

### Home ownership

```{r}
df |>
  group_by(person_home_ownership) |>
  mutate(freq = n()) |>
  ggplot() + 
  geom_bar(mapping=aes(x=reorder(person_home_ownership, desc(freq)))) +
  labs(title = "Home ownership modes for loan takers", x = "Home ownership mode")
```

### Loan intent

```{r}
df |>
  group_by(loan_intent) |>
  mutate(freq = n()) |>
  ggplot() +
    geom_bar(mapping = aes(x = reorder(loan_intent, desc(freq)))) + 
    labs(title = "Number of loans by intent" ,x = "Loan intent") +
    scale_x_discrete(guide = guide_axis(angle=60))
```

### Education level
```{r}
df |>
  group_by(person_education) |>
  mutate(freq = n()) |>
  ggplot() + 
    geom_bar(mapping = aes(x= reorder(person_education, desc(freq)))) +
    labs(title = "Education level for loan takers", x = "Education level")
```

### Gender

```{r}
df |>
  group_by(person_gender) |>
  mutate(freq = n()) |>
  ggplot() + 
    geom_bar(mapping = aes(x = reorder(person_gender, desc(freq)))) +
  labs(title = "Gender of loan takers", x = "Gender")
```
### Income

```{r}
ggplot(df) + 
  geom_histogram(mapping = aes(x = person_income), bins = 100) +
  labs(title = "Distribution of log of person income")
```

```{r}
ggplot(df) + 
  geom_histogram(mapping = aes(x = log(person_income)), bins = 100) +
  labs(title = "Distribution of log of person income")
```

## Bivariate Analysis

### Income based on education
```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = person_income, color = person_education)) + 
  labs(title = "Distribution of income based on education level", x = "Income")
```
```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = person_income, color = person_education), outliers=F, notch=T) +
  labs(title = "Distribution of income based on education level", x = "Income")
```
### Income based on gender
```{r}
ggplot(data = df) +
  geom_boxplot(mapping = aes(x = person_income, color = person_gender), outliers=F, notch=T) +
  labs(title = "Distribution of income based on gender", x = "Income")
```

### Loan amount based on income

```{r}
ggplot(df) + 
  geom_point(mapping = aes(x = person_income, y = loan_amnt), color="skyblue") +
  labs(title = "Loan amount based on income", x = "Income", y = "Loan Amount")
```

#### With log transformation on income

```{r}
ggplot(df) + 
  geom_point(mapping = aes(x = log(person_income), y = loan_amnt, color = person_gender), alpha = 0.2) +
  geom_smooth(mapping = aes(x = log(person_income), y = loan_amnt, color = person_gender), se = F) + 
  labs(title = "Loan amount vs log of income grouped by gender", x = "Log of income", y = "Loan amount")
```

#### With log transformation on both income and loan amount

```{r}
ggplot(data=df, mapping= aes(x=log(person_income), y=log(loan_amnt), color=loan_intent)) + 
  geom_point(alpha = 0.2) +
  geom_smooth(se = F) +
  labs(title = "Log of loan amount vs log of income grouped by loan intent", x = "log of income", y = "Log of loan amount")
```

# Modelling

## Linear Model using step-wise feature selection

```{r}
lm_min <- lm(person_income ~ 1, data = df)
lm_max <- lm(person_income ~ ., data = df)
```

```{r}
step(lm_min, scope = list(upper = lm_max), direction = "forward")
```

```{r}
step(lm_max, scope = list(lower = lm_min), direction = "backward")
```