---
title: "Loan Analysis"
author: "Collins Oduor"
date: "2026-03-31"
output: pdf_document
---

# Imports

```{r}
required_libraries <- c("dplyr",
                        "tidyverse",
                        "corrr",
                        "corrplot",
                        "MASS",
                        "DescTools")

for (required_library in required_libraries) {
  if(!(required_library %in% installed.packages())) {
    install.packages(required_library)
  }
  
  library(required_library, character.only = T)
}
```

# Configuration

```{r setup, include=TRUE}
# Ensure code blocks are shown on knitting to the target
knitr::opts_chunk$set(echo = TRUE)

# Centre all plot titles
theme_update(plot.title = element_text(hjust = 0.5))

# Remove plot background
theme_update(plot.background = element_blank())

# Remove panel background
theme_update(panel.background = element_blank())

fig.align = "center"
out.width = "90%"
```

# Loading data

```{r}
df <- read_csv(file="loan_data.csv")
```

# Exploratory Analysis
## Summary of data

```{r}
dim(df)
```
The data contains 45000 rows and 14 columns.

```{r}
head(df)
```

```{r}
names(df)
```

```{r}
sapply(df, anyNA)
```
All columns are 100% complete; no column has even a single null value.

## Univariate Analysis

### Home ownership

```{r}
df |>
  group_by(person_home_ownership) |>
  mutate(freq = n()) |>
  ggplot() + 
  geom_bar(mapping=aes(x=reorder(person_home_ownership, desc(freq)))) +
  labs(title = "Number of loan takers grouped by home ownership",
       x = "Home ownership mode",
       y = "Number of loan takers") +
  scale_y_continuous(limits = c(0, 25000)) +
  theme(panel.grid.major.y = element_line(color = "grey"),
        axis.ticks.y = element_blank())
```
Significantly fewer loans were taken by people who either own their homes or have a different ownership mode other than renting or mortgage. This could highlight the fact that having consistent monthly payments could squeeze one's budget to the extent that they are left with no means other than to take out a loan.

### Loan intent

```{r}
df |>
  group_by(loan_intent) |>
  mutate(freq = n()) |>
  ggplot() +
    geom_bar(mapping = aes(x = reorder(loan_intent, desc(freq)))) + 
    labs(title = "Number of loans by intent", x = "Loan intent") +
    scale_x_discrete(guide = guide_axis(angle=60)) +
    scale_y_continuous(name = "Number of applications", limits=c(0, 10000)) +
    theme(panel.grid.major.y = element_line(color = "grey"),
          axis.ticks.y = element_blank())
```
Fewer loans were taken for the purposes of home improvement than any other intent. Education and medical reasons were the most-cited reasons by loan takers.

### Education level

```{r}
df |>
  group_by(person_education) |>
  mutate(freq = n()) |>
  ggplot() + 
    geom_bar(mapping = aes(x= reorder(person_education, desc(freq)))) +
    labs(title = "Number of loan takers by education level",
         x = "Education level",
         y = "Number of loan takers") +
  scale_y_continuous(limits=c(0, 15000)) +
  theme(panel.grid.major.y = element_line("grey"),
        axis.ticks.y = element_blank())
```
Fewer loans were taken by people who hold at least a Master's degree. Interestingly, holders of Bachelor's degree took more loans than people who did not take their studies beyond high school.

### Gender

```{r}
df |>
  group_by(person_gender) |>
  summarise(freq = n(), .groups = "drop") |>
  ggplot(mapping = aes(x = "", y = freq, fill = person_gender)) + 
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    labs(title = "Composition of loan takers by gender") +
    theme_void() +
    geom_text(aes(label = person_gender),
              position = position_stack(vjust = 0.5)) +
    theme(legend.position = "none") +
    scale_fill_brewer(palette = "RdYlGn")
```

```{r}
df |>
  group_by(person_gender) |>
  mutate(freq = n()) |>
  ggplot() + 
    geom_bar(mapping = aes(x = reorder(person_gender, 
                                       desc(freq)))) +
  labs(title = "Number of loan takers by gender",
       x = "Gender",
       y = "Number of loan takers") + 
  theme(panel.grid.major.y = element_line("grey"),
        axis.ticks.y = element_blank())
```
There were marginally more male than female loan takers - about 5,000 more.

### Income

```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = person_income)) +
  theme(legend.position = "none") +
  labs(title = "Distribution of income", x = "Income") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
The distribution of income is skewed to the right. There are high earners who change the mean of the distribution.

```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = person_income), outliers=F) +
  theme(legend.position = "none") +
  labs(title = "Distribution of income (excluding outliers)",
       x = "Income") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```
Excluding outliers, it is clearer that the median income is about 70,000, and that 50% of the loan takers have an income of between (about) 50,000 and (about) 100,000. 

```{r}
ggplot(df) + 
  geom_histogram(mapping = aes(x = log(person_income)), bins = 100) +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_line("grey")) +
  scale_y_continuous(limits = c(0, 3000)) +
  geom_vline(mapping = aes(xintercept = median(log(df$person_income)))) +
  annotate("text",
           x = median(log(df$person_income))+0.7,
           y = 2945 ,
           label = paste0("Median: ",
                          format(round(
                            median(log(df$person_income)),
                            2),
                            nsmall = 2))) +
  labs(title = "Distribution of log of income",
       x = "Log of income") +
  theme(axis.ticks.y = element_blank(),
        axis.title.y = element_blank())
```
Log transformation was done on the income to narrow the range of possible values. The distribution of log of income now follows a fairly normal distribution.

### Age

```{r}
ggplot(df) + 
  geom_histogram(mapping = aes(x = person_age), bins = 50, boundary=0) +
  labs(title = "Distribution of age", x="Age") +
  theme(panel.grid.major.y = element_line("grey"),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```
The age of loan takers is skewed to the right.

```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = person_age), outliers=F) +
  theme(legend.position = "none") +
  labs(title = "Distribution of age (excluding outliers)",
       x = "Age") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```
As highlighted from the previous histogram, age is skewed to the right. Most of the loan takers have ages between 20 and just under 40. 

```{r}
nrow(subset(df, person_age > 40)) / nrow(df)
```

In fact, only 4% of the loan takers are above 40 years of age.

### Loan amount

```{r}
ggplot(df) + 
  geom_histogram(mapping = aes(x = loan_amnt), bins = 20) +
  labs(title = "Distribution of loan amount", x="Loan Amount") +
  theme(panel.grid.major.y = element_line("grey"),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  geom_vline(mapping = aes(xintercept = median(df$loan_amnt))) +
  annotate("text",
           x = median(df$loan_amnt) + 3700,
           y = 7800,
           label = paste0("Median: ", median(df$loan_amnt)))
```
The distribution of loan amount is skewed to the right, with a median of 8000.


### Employment experience

```{r}
ggplot(df) + 
  geom_histogram(mapping = aes(x = person_emp_exp)) +
  theme(legend.position = "none") +
  labs(title = "Distribution of employment experience",
       x = "Employment experience in years") +
  theme(axis.ticks.y = element_blank(),
        panel.grid.major.y = element_line("grey"),
        axis.title.y = element_blank()) +
  scale_y_continuous(limits = c(0, 20000)) +
  geom_vline(mapping = aes(xintercept = median(df$person_emp_exp))) +
  annotate("text",
           x = median(df$person_emp_exp) + 10,
           y = 19500,
           label = paste0("Median: ", median(df$person_emp_exp)))
```
Employment experience is skewed to the right, with a median of 4.

```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = person_emp_exp), outliers=F) +
  theme(legend.position = "none") +
  labs(title = "Distribution of employment experience (excluding outliers)",
       x = "Employment experience in years") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```
Most loan takers had employment experience of somewhere between 0 and just under 20 years.

```{r}
nrow(subset(df, person_emp_exp > 20)) / nrow(df)
```
Only 2% of loan takers had over 20 years of experience.

### Loan interest rate

```{r}
ggplot(df) + 
  geom_histogram(mapping = aes(x = loan_int_rate), bins = 20) +
  theme(legend.position = "none") +
  labs(title = "Distribution of loan interest rate",
       x = "Interest rate in percentage") +
  theme(axis.ticks.y = element_blank(),
        panel.grid.major.y = element_line("grey"),
        axis.title.y = element_blank()) +
  scale_y_continuous(limits = c(0, 10000)) +
  geom_vline(mapping = aes(xintercept = median(df$loan_int_rate))) +
  annotate("text",
           x = median(df$loan_int_rate) + 1.5,
           y = 9500,
           label = paste0("Median: ", median(df$loan_int_rate)))
```
The distribution of interest rates follows a fairly normal distribution, with a median of about 11%.


### Loan percent income

```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = loan_percent_income)) +
  theme(legend.position = "none") +
  labs(title = "Distribution of loan income",
       x = "Loan income in percentage") +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank())
```
Most loans yielded an income of between 0% to 0.4%. There were a few outliers that yielded a profit of over 0.4%.

### Previous loan default

```{r}
df |>
  group_by(previous_loan_defaults_on_file) |>
  summarise(freq = n(), .groups = "drop") |>
  ggplot(mapping = aes(x = "", y = freq,
                       fill = previous_loan_defaults_on_file)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  geom_text(mapping = aes(label = previous_loan_defaults_on_file),
            position = position_stack(vjust = 0.5)) +
  theme(legend.position = "none") +
  ggtitle(label = "Previous default rates",
          subtitle = "As many people have defaulted previously as have not")
```
Previous default rate for loan takers is about 50% - as many applicants have previously defaulted as have not defaulted.


### Credit score

```{r}
ggplot(df) +
  geom_histogram(mapping = aes(x = credit_score), bins = 50) +
  theme(axis.title.y = element_blank(),
        panel.grid.major.y = element_line("grey"),
        axis.ticks.y = element_blank()) + 
  scale_y_continuous(limits = c(0, 4000)) +
  labs(title = "Distribution of credit score", x = "Credit score")
```

```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = credit_score)) +
  theme(legend.position = "none") +
  labs(title = "Distribution of credit score",
       x = "Credit score") +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank())
```

The distribution credit score appears to be slightly skewed to the left, with most loan takers having scores between 500 and 800. Outliers were observed at both tails of the distribution (with more outliers towards the left tail).

### Loan status

```{r}
df |>
  group_by(loan_status) |>
  summarise(freq = n(), .groups = "drop") |>
  ggplot(mapping = aes(x="", y = freq, fill = as.factor(loan_status))) +
  geom_bar(stat = "identity", width = 1) + 
  coord_polar("y", start = 0) + 
  theme_void() +
  geom_text(mapping = aes(label = ifelse(loan_status,
                                         "Repaid",
                                         "Defaulted")),
            position = position_stack(vjust = 0.5)) +
  theme(legend.position = "none") +
  ggtitle("Analysis of loan perfomance",
          subtitle = "More than 75% of loans were defaulted")
```
Most of the loans (>75%) did not perform in the sense that they were eventually defaulted.

## Bivariate Analysis

### Income

#### Based on education

```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = person_income,
                             color = person_education)) + 
  labs(title = "Distribution of income based on education level",
       x = "Income") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
Even though outliers in terms of income occur at all education levels, the outliers are less pronounced for holders of Masters and Doctorate degrees. 


```{r}
ggplot(df) + 
  geom_boxplot(mapping = aes(x = person_income,
                             color = person_education),
               outliers=F,
               notch=T) +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  labs(title = "Doctorate degree holders have a higher median income than others",
       x = "Income")
```
Ignoring outliers, it becomes clear that the distribution of income for Doctorate degree holders has a rightward shift relative to people at other education levels. The notch of the boxplot not overlapping with other boxplots indicates that the median is significantly different.

#### Based on gender

```{r}
ggplot(df) + 
  geom_histogram(mapping = aes(x = person_income, fill = person_gender), bins = 30) +
  xlim(0, 1.5e+05) +
  facet_grid(person_gender ~ .) +
  theme(legend.position = "none") +
  theme(panel.grid.major.y = element_line("grey")) +
  scale_y_continuous(limits = c(0, 2000)) +
  labs(title = "The distributions of income both genders greatly overlap",
       x = "Income",
       y = "Number of people at income level")
```
The distribution of income by gender shows an almost-perfect overlap, indicating that there is no difference in income based on gender.

#### Based on home ownership

```{r}
ggplot(df) + 
  geom_histogram(mapping = aes(x = person_income, fill = person_home_ownership)) +
  xlim(0, 1.5e+05) +
  facet_grid(person_home_ownership ~ .) +
  theme(legend.position = "none") +
  theme(panel.grid.major.y = element_line("grey")) +
  scale_y_continuous(limits = c(0, 2000)) +
  labs(title = "Distributions of income based on home ownership",
       x = "Income",
       y = "Number of people at income level")
```
The distributions of income for people who pay neither rent nor mortgage is essentially non-existent on the scale. This is due to how few loans were taken by those who do not have home payments.

#### Based on years of experience

```{r}
ggplot(data = df) +
  geom_point(mapping = aes(x = person_emp_exp,
                           y = person_income,
                           color = person_education)) +
  ylim(0, 1.5e+05) +
  xlim(0, 40) +
  labs(title = "Income by years of experience",
       y = "Income",
       x = "Years of Experience")
```
There does not appear to be any relationship between the income and experience, even accounting for level of education.

### Loan amount

#### Based on income

```{r}
ggplot(df) + 
  geom_point(mapping = aes(x = person_income, y = loan_amnt), color="skyblue") +
  labs(title = "Loan amount based on income", x = "Income", y = "Loan Amount") +
  annotate("text", label = "Elite earners", x=6.3e+06, y = 5000)
```
Interestingly, the elites who out-earn everyone else were not the takers of the largest loans.

```{r}
ggplot(df) + 
  geom_point(mapping = aes(x = person_income, y = loan_amnt), color="skyblue") +
  labs(title = "Loan amount based on income", x = "Income", y = "Loan Amount") +
  xlim(0, 1.5e+05)
```
Accounting for outliers in income, there does not appear to be any relationship between loan amount and income.

##### With log transformation on income

```{r}
ggplot(df) + 
  geom_point(mapping = aes(x = log(person_income),
                           y = loan_amnt,
                           color = person_gender),
             alpha = 0.2) +
  geom_smooth(mapping = aes(x = log(person_income),
                            y = loan_amnt,
                            color = person_gender),
              se = F) + 
  labs(title = "Loan amount vs log of income grouped by gender",
       x = "Log of income",
       y = "Loan amount")
```

##### With log transformation on both income and loan amount

```{r}
ggplot(data=df, mapping= aes(x=log(person_income),
                             y=log(loan_amnt),
                             color=loan_intent)) + 
  geom_point(alpha = 0.2) +
  geom_smooth(se = F) +
  labs(title = "Log of loan amount vs log of income grouped by loan intent",
       x = "log of income",
       y = "Log of loan amount")
```
 Transforming the loan amount and income variables using log did not increase the level of relationship between the 2 variables.
 
 
### Loan interest rate

#### Based on loan amount

```{r}
ggplot(data = df, mapping = aes(x = loan_amnt,
                                y = loan_int_rate,
                                color = factor(loan_status))) +
  geom_point() + 
  labs(title = "Loan interest rate based on loan amount",
       x = "Loan amount",
       y = "Loan interest rate",
       fill = "Loan repayment s1tatus")
```
 
### Loan Status

#### Based on gender

```{r}
ggplot(data = df, mapping = aes(x = person_gender,
                                fill = factor(loan_status))) +
  geom_bar(position = "dodge") +
  scale_fill_discrete(labels = c("0" = "Defaulted", "1" = "Repaid")) +
  labs(title = "Loan repayment based on gender",
       x = "Gender",
      y = "Count of loans",
      fill = "Loan Status")
```
The default rate for male loan takers is higher than that or females. This is seen from the difference in relative heights of the bar representing defaulted and repaid loans for either gender.

#### Based on education

```{r}
ggplot(data = df, mapping = aes(x = person_education,
                                fill = factor(loan_status))) +
  geom_bar(position = "dodge") +
  scale_fill_discrete(labels = c("0" = "Defaulted", "1" = "Repaid")) +
  labs(title = "Loan repayment based on education level",
       x = "Education level",
      y = "Count of loans",
      fill = "Loan Status") +
  scale_x_discrete(limits = c("High School",
                              "Associate",
                              "Bachelor",
                              "Master",
                              "Doctorate"))
```
Doctorate degre holders take the fewest number loans, and have the fewest defaults of any education level.

## Correlation Analysis

```{r}
correlation_matrix <- cor(df[, sapply(df, is.numeric)])
corrplot(correlation_matrix,
         method = "shade",
         type = "lower",
         order = "hclust",
         tl.col = "black",
         tl.srt = 45)
```

# Feature Engineering

```{r}
df_clean = df |>
  mutate(person_age = Winsorize(person_age, val=c(20, 40))) |>
  mutate(person_emp_exp = Winsorize(person_emp_exp, val = c(0, 20))) |>
  mutate(person_income = Winsorize(person_income, val = c(0, 150000)))
```


# Modelling

I would like to answer the following questions:

1. Can one's income be predicted using the other features?
2. Can the loan repayment status be predicted given the other features?

## Model to predict income

### Linear Model using step-wise feature selection

```{r}
lm_min <- lm(person_income ~ 1, data = df_clean)
lm_max <- lm(person_income ~ ., data = df_clean)
```

```{r}
step(lm_min, scope = list(upper = lm_max), direction = "forward")
```

```{r}
step(lm_max, scope = list(lower = lm_min), direction = "backward")
```

## Model to predict loan repayment status

